---
title: "Simple xarray example in R"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document presents an example of how the [xarray](https://docs.xarray.dev/en/stable/index.html) Python library can be used in R.

## Installation

To use the [xarray](https://docs.xarray.dev/en/stable/index.html) in R, it is required to have the [reticulate](https://rstudio.github.io/reticulate/) package installed. To do so, you can use the following command:

```{r, eval=FALSE}
install.packages("reticulate")
```

After installing the [reticulate](https://rstudio.github.io/reticulate/) package, we must create a Python virtual environment to install the [xarray](https://docs.xarray.dev/en/stable/index.html) library. For this, we can use the following command:

> In this example, we use [conda](https://docs.conda.io/en/latest/) to manage Python virtual environments, but any other package management tool can also be used (e.g., mamba, virtualenv, poetry).

```{r}
# defining the name of the environment
conda_env_name <- "r-xarray"

# creating conda environment
reticulate::conda_create(envname = conda_env_name)
```

With the environment created, let's install the dependencies on it:

```{r}
reticulate::py_install(
  c("xarray"),
  envname = conda_env_name,
  pip =  TRUE
)
```

## Importing Python libraries

Now, we have an environment ready to go. So, let's activate it:

```{r}
reticulate::use_condaenv(condaenv = conda_env_name)
```

To use the [xarray](https://docs.xarray.dev/en/stable/index.html) library, we can import it using the following [reticulate](https://rstudio.github.io/reticulate/) command:

```{r}
xr <- reticulate::import("xarray")
```

For this example, we will also use other Python libraries. So, we also need to import them:

> As you can see in the code below, we are importing [pandas](https://pandas.pydata.org/) and [numpy](https://numpy.org/). Both are dependencies from [xarray](https://docs.xarray.dev/en/stable/index.html), so we don't need to install them directly.

```{r}
pd <- reticulate::import("pandas")
np <- reticulate::import("numpy")
```

## Dummy dataset example

To show how the [xarray](https://docs.xarray.dev/en/stable/index.html) can handle multidimensional data, we will create a dummy dataset with 3 dimensions: `X`, `Y`, and `Y`.

For this, let's create a 3D array using [numpy](https://numpy.org/) and random values:

```{r}
data <- np$random$rand(1000L, 100L, 10L)
```

We already have the data created. Now, we need to define the name of its dimensions:

```{r}
dims <- c("x", "y", "time")
dims
```

As the [xarray](https://docs.xarray.dev/en/stable/index.html) uses indices to handle data, we also need to specify the coordinates of the dataset:

> The `coordinates` are used to specify the extensions of the data (e.g., spatial dimension (`X`, `Y`) and temporal dimension (`Y`))

```{r}
coords = list(
  x = 1:1000,
  y = 1:100,
  time = pd$period_range("2000-01-01", periods = 10)
)
```

Using the elements we have defined in the code blocks above, let's create a [DataArray](https://docs.xarray.dev/en/stable/generated/xarray.DataArray.html):

```{r}
ds <- xr$DataArray(data, dims = dims, coords = coords)
ds
```

## Temporal mean

With the [DataArray](https://docs.xarray.dev/en/stable/generated/xarray.DataArray.html) created, it can be manipulated (e.g., extract values, calculate statistics). As an example, let's calculate the temporal mean:

```{r}
ds_mean <- ds$mean(dim = "time")
ds_mean
```

By using the `mean(dim = "time")`, we specify to the xarray the mean should be calculated in the temporal dimension. So, in our example, it is calculated using `10 values` values for each `X` and `Y`.
