---
title: "Using xarray with dask"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document presents an example of how the [xarray](https://docs.xarray.dev/en/stable/index.html) and the [dask](https://www.dask.org/) Python libraries can be used together in R.

## Installation

To use the [xarray](https://docs.xarray.dev/en/stable/index.html) in R, it is required to have the [reticulate](https://rstudio.github.io/reticulate/) package installed. To do so, you can use the following command:

```{r, eval=FALSE}
install.packages("reticulate")
```

After installing the [reticulate](https://rstudio.github.io/reticulate/) package, we must create a Python virtual environment to install the [xarray](https://docs.xarray.dev/en/stable/index.html) library. For this, we can use the following command:

> In this example, we use [conda](https://docs.conda.io/en/latest/) to manage Python virtual environments, but any other package management tool can also be used (e.g., mamba, virtualenv, poetry).

```{r}
# defining the name of the environment
conda_env_name <- "r-xarray"

# creating conda environment
reticulate::conda_create(envname = conda_env_name)
```

With the environment created, let's install the dependencies on it:

```{r}
reticulate::py_install(
  c("xarray", "dask", "distributed"),
  envname = conda_env_name,
  pip =  TRUE
)
```

## Importing Python libraries

Now, we have an environment ready to go. So, let's activate it:

```{r}
reticulate::use_condaenv(condaenv = conda_env_name)
```

To use the [xarray](https://docs.xarray.dev/en/stable/index.html) library, we can import it using the following [reticulate](https://rstudio.github.io/reticulate/) command:

```{r}
xr <- reticulate::import("xarray")
```

For this example, we also need to import some [dask](https://www.dask.org/) packages:

```{r}
da <- reticulate::import("dask.array")
dd <- reticulate::import("dask.distributed")
```

To create some dummy data, we must also import [pandas](https://pandas.pydata.org/):

```{r}
pd <- reticulate::import("pandas")
```

## Configuring Dask

As [dask](https://www.dask.org/) was created to help us scale applications using multiple cores or machines, to use it, we need first to configure the connection between our application and those resources. To do that, we must create an instance of the class `dask.distributed.Client`:

```{r}
client <- dd$Client()
client
```

By default, when we create an instance of the `dask.distributed.Client` class, [dask](https://www.dask.org/) configures a local server to use all cores available in the machine.

## Dummy dataset example

To show how the [xarray](https://docs.xarray.dev/en/stable/index.html) can use parallel-enabled arrays from [dask](https://www.dask.org/), we will create a dummy dataset with 3 dimensions: `X`, `Y`, and `Y`.

> Using dask array is possible because xarray can work with multiple "array backends". To learn more, please check this [documentation page](https://docs.xarray.dev/en/stable/user-guide/duckarrays.html)

For this, let's create a 3D array using [Dask array](https://docs.dask.org/en/stable/array.html) with values:

```{r}
data <- da$random$random(c(10000, 1000, 100), chunks = c(1000, 1000, 10))
```

We already have the data created. Now, we need to define the name of its dimensions:

```{r}
dims <- c("x", "y", "time")
dims
```

As the [xarray](https://docs.xarray.dev/en/stable/index.html) uses indices to handle data, we also need to specify the coordinates of the dataset:

> The `coordinates` are used to specify the extensions of the data (e.g., spatial dimension (`X`, `Y`) and temporal dimension (`Y`))

```{r}
coords = list(
  x = 1:10000,
  y = 1:1000,
  time = pd$period_range("2000-01-01", periods = 100)
)
```

Using the elements we have defined in the code blocks above, let's create a [DataArray](https://docs.xarray.dev/en/stable/generated/xarray.DataArray.html):

```{r}
ds <- xr$DataArray(data, dims = dims, coords = coords)
ds
```

## Temporal mean

With the [DataArray](https://docs.xarray.dev/en/stable/generated/xarray.DataArray.html) created, it can be manipulated (e.g., extract values, calculate statistics). As an example, let's calculate the temporal mean:

```{r}
ds_mean <- ds$mean(dim = "time")
ds_mean
```

By using the `mean(dim = "time")`, we specify to the [xarray](https://docs.xarray.dev/en/stable/index.html) the mean should be calculated in the temporal dimension. So, in our example, it is calculated using `10 values` values for each `X` and `Y`.

With [xarray](https://docs.xarray.dev/en/stable/index.html), we are always using lazy evaluation functions. So, the code above still needs to be executed. For this, we can ask [dask](https://www.dask.org/) to execute it using the `compute` method.

```{r}
ds_mean_result <- ds_mean$compute()
ds_mean_result
```
